\hypertarget{preface}{%
\subsection{PREFACE}\label{preface}}

This document is intended to introduce pointers to beginning programmers
in the C programming language. Over several years of reading and
contributing to various conferences on C including those on the FidoNet
and UseNet, I have noted a large number of newcomers to C appear to have
a difficult time in grasping the fundamentals of pointers. I therefore
undertook the task of trying to explain them in plain language with lots
of examples.

The first version of this document was placed in the public domain, as
is this one. It was picked up by Bob Stout who included it as a file
called PTR-HELP.TXT in his widely distributed collection of SNIPPETS.
Since that original 1995 release, I have added a significant amount of
material and made some minor corrections in the original work.

In the HTML version 1.1 I made a number of minor changes to the wording
as a result of comments emailed to me from around the world.~ In version
1.2 I updated the first two chapters to acknowledge the shift from 16
bit compilers to 32 bit compilers on PCs.

In version 1.3, there is a Markdown version of the turorial available.

\hypertarget{acknowledgements}{%
\subsubsection{Acknowledgements:}\label{acknowledgements}}

There are so many people who have unknowingly contributed to this work
because of the questions they have posed in the FidoNet C Echo, or the
UseNet Newsgroup comp.lang.c, or several other conferences in other
networks, that it would be impossible to list them all. Special thanks
go to Bob Stout who was kind enough to include the first version of this
material in his SNIPPETS file.

\hypertarget{about-the-author}{%
\subsubsection{About the Author:}\label{about-the-author}}

Ted Jensen is a retired Electronics Engineer who worked as a hardware
designer or manager of hardware designers in the field of magnetic
recording. Programming has been a hobby of his off and on since 1968
when he learned how to keypunch cards for submission to be run on a
mainframe. (The mainframe had 64K of magnetic core memory!).

\hypertarget{about-the-maintainer}{%
\subsubsection{About the Maintainer}\label{about-the-maintainer}}

Jay Flaherty is a Principal Software Engineer who has worked for various
entities programming in everything from Perl to Python to Java and now
is diving deep into the world of C programming.

Use of this Material:

\hypertarget{other-versions-of-this-document}{%
\subsubsection{Other versions of this
document:}\label{other-versions-of-this-document}}

The original 1.2 version of this document is available at
\href{https://github.com/jflaherty/ptrtut13/releases}{Github Releases}

\begin{comment}
\href{introx.htm}{Introduction}

\href{pointers.htm}{Back to Table of Contents}
\end{comment}
